Nuclear Waste Management and Disposal

### 2017-03-26 paper.tex
Alex's comments

%\documentclass{article}
\documentclass[nofootinbib,preprint,aps]{revtex4-1}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{xspace}
\usepackage{csquotes}
\usepackage{gensymb}
\usepackage{graphicx}
\graphicspath{ {images/} }
\usepackage{physics}
\usepackage{color}
\newcommand{\red}[1]{\textcolor{red}{\bf #1}}
\begin{document}
\title{Nuclear Waste Management and Disposal}

@@

\date{\today}%
\begin{abstract}
    As nuclear reactors around the world continuously generate highly radioactive waste,
    how we manage nuclear waste is becoming more pertinent. In this work,
    we examine a few scientific and technical issues related to nuclear waste management and disposal.
    The principles of radioactivity and nuclear fission help us to understand the sources and properties
    of different types of nuclear waste. Focusing on the highly radioactive spent fuel, the rest
    of this work outlines the advantages and drawbacks of different spent fuel management and disposal
    methods. We study the methods used for performance assessments for Yucca Mountain Nuclear Waste Repository,
    especially in the context of volcanic disruption. Finally, we discuss some possible options for 
    nuclear waste management and disposal after
    the termination of the Yucca Mountain project.\\
    Word count: 7449\\
\end{abstract}
\maketitle
\tableofcontents
\newpage
\section{Introduction}
Nuclear power continues to play an important role in global energy supply.
The International Atomic Energy Agency (IAEA) estimates that nuclear power
consumption will increase by $40\%$ by 2030.\cite{iaea12}
Whereas the ``front end'' of the nuclear fuel cycle, such as reactor physics and power output, draws
the most attention from the public, the ``back end'' of the fuel cycle, i.e. waste management and
disposal, deserves equal, if not more attention.
Currently, the IAEA estimates that there are over 10,000 metric tons of highly radioactive nuclear waste 
generated worldwide annually,
most of which is spent nuclear fuel from nuclear reactors.\cite{iaea08, r12}
The world will have over $457,000$ metric tons of highly radioactive waste stock by the end of 2020.\cite{r12}
Highly radioactive waste remains hazardous for up to $10^6$ years after it is produced and thus
requires extremely careful disposal methods.
Moreover, there is currently no permanent repository in the world to dispose of these highly radioactive
waste.\footnote{The Waste Isolation Pilot Plant (WIPP) in New Mexico is designed to be a permanent radioactive
    waste repository, but only
for waste classified as transuranic, which, by the Nuclear Regulatory Commission standard, does not have particularly
high radioactivity but only long half-life (See Sec.~\ref{sec:waste:classification}). In particular, WIPP does not accept spent nuclear
fuel.}

It is alarming that more than $75\%$ of the spent nuclear fuel worldwide is stored on-site at the nuclear power plants
in water pools that have exceeded their designed capacity and are reaching their limiting capacity.\cite{aa12}
Two weeks after the devastating earthquake and tsunami hit the Fukushima Daiichi Nuclear Power Plant in Japan,
the water pool cooling system was cut off from power. Officials feared that
the highly radioactive waste in the pools would catch fire and send radioactive smoke
across a much larger part of Japan, including Tokyo.\cite{s16} For some ``fortuitous'' reasons, the fear did not come true.
However, the Nuclear Regulatory Commission (NRC), the independent government agency
of the U.S. government overseeing nuclear energy safety,
estimates that a major fire in a U.S. spent fuel pool is expected to
displace 3.4 million people in an area larger than New Jersey.\cite{s16}

@@

In this work, we will attempt to examine some of the scientific issues underlying nuclear waste management
and disposal. We will start by introducing radiations and its biological effects (Section~\ref{sec:phys}); then
we examine the sources and the physical properties of nuclear waste (Section~\ref{sec:waste}),
its management and temporary storage (Section~\ref{sec:temp}), and the debates around its eventual disposal (Section~\ref{sec:disposal}).

\section{Radioactivity and Its Biological Effects}
\label{sec:phys}
The principles of radioactivity and its biological effects play a central role in our
discussion of nuclear waste. Hence, we should first review and extend our understanding of 
radioactivity and radiations.
    \subsection{Radioactivity\footnote{
    The derivations and in this section follows chapter $6$ in
    Krane's text \cite{k88} and chapter $1$ in Lilley's text \cite{l01}.
    }}
    Radioactive decay is the process in which most of the known detrimental radiations are generated.
    Here we review the physical principles and mathematical formalism behind it.
        \subsubsection{Quantum theory of radioactive decays}
        The law of radioactive decay is a direct consequence of quantum mechanics.
        Consider a nucleus going through radioactive decay.
        We can approximate the nucleus as a particle situated in a potential well with potential given by
        \begin{equation}
            \label{eq:vp}
            \tilde{V}(t) = V + V_p,
        \end{equation}
        where $V$ is the time-independent nucleus potential that gives rise to eigenstates $\ket{\psi_1}$
        and $\ket{\psi_2}$ with $\ket{\psi_2}$ having a lower (expected) energy; $V_p$ is a small
        perturbation that causes the decay.
        Suppose the particle is in a state given by
        \begin{equation}
            \label{eq:state}
            \ket{\psi(t)} = c_1(t)\ket{\psi_1} + c_2(t)\ket{\psi_2},
        \end{equation}
        where $c_1,c_2$ are the coefficients of superposition.
        A decay from $\ket{\psi_1}$ to $\ket{\psi_2}$ is equivalent to
        having $|c_2(t)|^2\gg |c_1(t)|^2$ as time increases. Since we know that time-independent potentials
        only give rise to time-independent states, the fact that Eq.~\ref{eq:state} has time dependence requires
        our perturbation $V_p$ in Eq.~\ref{eq:vp} to be time dependent.

@@

        Specific activity, on the other hand, gives the activity per mass of a sample
        \begin{equation}
        \mathcal{A}_s = \frac{\mathcal{A}(0)}{{m_{sample}}}.
        \end{equation}
        Specific activity can characterize the activity level of a certain isotope or compound.
        The difference between
        activity and specific activity resembles the difference between heat capacity and specific heat
        capacity.
        The radioactive isotope used for Computer Tomography (CT) scan has a specific activity
        of $\sim 0.01$ Ci/g. See Table~\ref{tab:hl} for the half-lives and specific activities
        of some isotopes present in highly radioactive nuclear waste.
        \begin{table}
            \centering
            \caption{Specific activity and half-life of isotopes commonly present in spent nuclear fuel \cite{toi}}
            \begin{ruledtabular}
                \begin{tabular}{l l l}
                Isotope & Half-life (Years) & Specific Activity (Ci/g)\\
                \hline
                Cesium-137 & $37$ & $88$\\
                Thorium-232 & $1.4 \times 10^{10}$ & $1.1\times 10^{-7}$\\
                Uranium-233 & $1.6\times 10^6$ & $9.3\times 10^{-3}$\\
                Plutonium-239 & $2.4\times 10^4$ & $0.063$\\
                Neptunium-237 & $2.1\times 10^6$ & $6.9\times 10^{-4}$\\
            \end{tabular}
            \label{tab:hl}
            \end{ruledtabular}
        \end{table}
        \subsubsection{Types of radiation}
        There are three main types of radioactive decay processes that generate radiations in nuclear waste: 
        Alpha decay, which generates alpha particles (helium-4 nuclei), beta minus decay which generates beta
        particles (electrons) and gamma decay which generates gamma rays.
        The type of radiation (particle beam) generated
        have different properties and deserve separate treatments when we discuss health effects and shielding.
        \paragraph{Alpha decay}
        An alpha particle is a helium-4 atom ${}^4_2$He. In an alpha decay, a nucleus emits an alpha
        particle and decays into a different element. The emission of an alpha particle is
        particularly favorable because it has a particularly
        high binding energy. The process can be represented by
        \begin{equation}
            {}^A_Z X_N \rightarrow {}^{A-4}_{Z-2}X'_{N-2} + \alpha,
        \end{equation}
        where $Z$ denotes the atomic number (number of protons) of a nucleus $A$ the atomic mass, and $N$ the
        neutron count.

@@

        \paragraph{Gamma decay}
        Most nuclear reactions, including the alpha and beta decay, leave the resulting nucleus in
        an excited state.
        We know that excited atom radiates photons to fall back into ground state. Similarly, excited nuclei
        decay to ground state via emitting photons, with energies in the range
        of $0.1$ to $10$ MeV, which are characteristic of the energy differences between nuclear energy states.

        \subsubsection{Daughter nuclei and subsequent decay\footnote{The derivation in this section follows that of
        \cite[chapt. 1.5]{l01}}}
        The resultant nuclei (called ``daughter products'') from a radioactive decay are often radioactive
        themselves and their decay rate can be larger or smaller than their parents. Consider a decay chain
        $A\rightarrow B \rightarrow C \rightarrow ...$ with decay constants $\lambda_A, \lambda_B,...$. Suppose
        we start with a sample of $N_0$ $A$ nuclei. The change of the number of nuclei $B$ increases as $A$
        decays and decreases s $B$ decay. Hence we have
        \begin{equation}
            \frac{dN_B}{dt} = -\lambda_B N_B + \lambda_A N_A.
        \end{equation}
        Multiply both sides by $e^{\lambda_B t}$ and substitute in $N_A=N_0 e^{-\lambda_A} t$ to get
        \begin{align*}
            \frac{dN_B}{dt}e^{\lambda_B t} + \lambda_B N_B e^{\lambda_B t} &= \lambda_A N_A e^{\lambda_B},\nonumber\\
            \frac{d}{dt} (N_B e^{\lambda_B t}) &= \lambda_A N_0 e^{(\lambda_B-\lambda_A)t}, \nonumber\\
            N_B e^{\lambda_B t} &= \lambda_A N_0 \int\limits_0^t e^{(\lambda_B - \lambda_A)u} du \nonumber\\
                                &= \frac{\lambda_A}{\lambda_B-\lambda_A}N_0 e^{(\lambda_B - \lambda_A)t}
        \end{align*}
        Hence we end up with the number of $B$ nuclei as a function of time
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B-\lambda_A}N_0(e^{-\lambda_A t} - e^{-\lambda_B t}).
            \label{eq:daughter}
        \end{equation}
        Similar derivations can be done for the rest of the isotopes in the decay chain by considering
        one decay in a chain at a time. If the daughter isotope has a longer half-life than the parent
        ($\lambda_A \gg \lambda_B$), for small $t$, Eq.~\ref{eq:daughter} becomes
        $N_B(t)\sim N_0 e^{-\lambda_A t}$. After a long time,
        Eq.~\ref{eq:daughter} becomes $N_B(t) \sim N_0 e^{-\lambda_B t}$; the decay
        chain can be treated as the fast decay of $A$ into $B$, and then the slow decay of $B$
        into $C$.

        On the other hand, if the parent is longer lived
        ($\lambda_B \gg \lambda_A)$, after a long time, Eq.~\ref{eq:daughter} can be written as
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B}N_0 e^{-\lambda_A t}=\frac{\lambda_A}{\lambda_B}N_A(t).
        \end{equation}
        Hence the number of nuclei $A$ and nuclei $B$ are in equilibrium determined by their decay rates, with
        $A$ dominating the nuclei count. Taking the limit as $\lambda_A \rightarrow 0$, $N_B(t)\rightarrow 0$.
        The decay is essentially the decay of $A$ directly into $C$. Hence, if we have a chain of reaction
        where each daughter nucleus has a half-life much shorter than that of its parent, then 
        isotope $A$ effectively decays into the end of the decay chain and releases a lot more radiations than
        in a single decay. Moreover, before the decay chain reaches its equilibrium, the activity of the sample
        may increase.

        \subsection{Biological effects of radiation\footnote{Much of the discussion in this section follows that in
        Lilley's text \cite[chapt. 7]{l01}}}
        The biological effects of radiation have been an important topic for decades. A large
        number of studies took place after the two nuclear bomb explosions in Japan in 1945
        and after disastrous nuclear plant accidents like Chernobyl. Therefore, we have a relatively
        thorough understanding of the consequences of mass exposure to radiation. However, relatively
        little is known about the long-term effects of exposure to small doses, including subtle biological
        changes that lead to cancer and genetic defects. Here we briefly summarize the known effects.
        \subsubsection{Interactions of radiations with living tissues}
        \label{sec:interactions}
        Radiations causes damage to living tissues by depositing energy into the biological molecules.
        Depending on the type of the radiation
        and the molecules present in the tissues, the damage mechanism and effects are different.
        \paragraph{Alpha particles} An alpha particle loses energy predominantly via the interactions between its electric
        field and the electrons in the matter it travels through. Interactions with nuclei and head-on collisions
        are quite rare. Since the speed of an alpha particle is low, it deposits more energy as it
        travels. In tissues, $1$-MeV alpha particles travel through only a few tens of micrometers and thus
        are easily stopped by skin. However, if alpha particles are ingested and more sensitive organs are irradiated,
        serious damages can be done.
        \paragraph{Beta radiation (elections)} Electrons are much lighter and hence are easily scattered.
        Moreover, they are much more penetrating since their energy lost per distance traveled is lower.
        Finally, electrons also radiate photons
        via bremsstrahlung radiations as they decelerate. Therefore, the electrons' energy end up distributing
        across a much larger volume.
        \paragraph{Photons} Gamma rays interact with matter via photoelectric effects, Compton scattering and pair productions
        depending on the energy of the photon and the atoms it interact with.
        Since biological tissues mainly consist of elements of low atomic number, Compton
        scattering is the most important interaction. The scattered photons may interact further with the tissues
        via photoelectric effects and secondary Compton scattering. As a result, the intensity of the energy deposition falls
        off much more slowly and is hence more penetrating. Both electrons and photons can penetrate human body easily.
        Although they do not deposit as much energy as alpha particles, they are generally more dangerous due to their
        penetration abilities.

        The energy deposition described above is an important cause of irreversible damage to sensitive molecules
        like the Deoxyribonucleic (DNA) molecules if the energy transferred is large enough.
        In the cases where the energy transferred is not as large, the radiation ionizes simpler molecules like water
        to form chemically active free radicals. The free radicals have unpaired electrons. Hence they go on to react
        and disrupt
        other molecules that may be crucial to the biological system.

        \subsubsection{Quantifying radiation on living tissues}
        The biological impact of radiation depends on the type, power and energy of the radiation, the
        exposure time, and the part of the body exposed. Various quantities have been introduced to
        the biological impact of radiation. Here we review three
        of these quantities: the absorbed dose, the equivalent dose and the effective dose.
        The SI units for dose is sievert (Sv), which is equivalent to J/Kg, the amount of energy deposited into
        per unit mass of living tissue. Some work uses the unit gray (Gy) for absorbed dose in medical context,
        but it is the same as sievert.

        The effects of radiation is usually characterized by the amount of energy it carries since it constrains
        the amount of energy it deposits in living tissues.
        Absorbed dose~($D$) quantifies the 
        radiation energy absorbed per unit mass.

        For an external $\gamma$-ray source emitting $\mathcal{A}$ million photons per second 
        of energy $E_{\gamma}$
        (in MeV), one can approximate the dose rate (dose received per unit of time) by \cite[chapt. 7]{l01}
        \begin{equation}
            \frac{dD}{dt}(\mu\text{Sv h}^{-1}) \approx \frac{\mathcal{A}\text{(MBq)}\times E_{\gamma}\text{(MeV)}}
            {6\times [r\text{(m)}]^2}.
        \end{equation}
        The approximate linear dependence on $E_{\gamma}$ is valid from less than $0.1$ MeV to several
        MeV and therefore should be good for all sources we are considering.\cite{my68}
        The inverse-squared dependence comes from the isotropic radiation assumption.

        \begin{table}[h]
            \centering
            \caption{Weighting factors for different types of radiation \cite{icrp74}}
            \begin{ruledtabular}
                \begin{tabular}{l c c}
                Type of radiation & Energy range & Weighting factor, $w_R$\\
                \hline
                Photons, electrons & All energies & 1\\
                Neutrons & $<10$ keV & 5 \\
                         & $10-100$ keV & 10 \\
                         & $100$ keV-$2$ MeV & 20 \\
                         & $2-20$ MeV & 10 \\
                Protons & $<20$ MeV & 5 \\
                Alpha particles, fission fragments, heavy nuclei & & $20$\\
            \end{tabular}
            \label{tab:eq}
            \end{ruledtabular}
        \end{table}
        However, absorbed dose does not take into account different biological implications of different types
        of radiation.
        For example, the same dose of alpha particle deposit more energy than gamma rays. The equivalent
        dose ($H$) is defined as
        \begin{equation}
        H=w_R \times D_R,
        \end{equation}
        where $D_R$ is the absorbed dose and $w_R$ the weighting factor of a given kind of radiation $R$
        (see Table~\ref{tab:eq} for a list of weighing factors).

        Finally, certain organs and parts of the body are more vulnerable to radiation than the rest.
        Therefore, effective dose~($E$) is defined as the equivalent dose weighted by the body part
        radiated
        \begin{equation}
            E = w_T \times H_T,
        \end{equation}
        where $w_T$ is the body tissue weighting factor, and $H_T$ the equivalent dose on the tissue
        at question.
        Some of the recommended values of tissue
        weighting factors are given in Table~\ref{tab:eff}. A weighing factor of $1$ is equivalent to the
        whole body absorbing the radiation uniformly.
        \begin{table}[h]
            \centering
            \caption{Weighting factors for individual organs \cite{icrp74}}
            \begin{ruledtabular}
                \begin{tabular}{l c}
                Tissue & Weighting factor, $w_T$\\
                \hline
                Gonads & $0.20$\\
                Red bone marrow & $0.12$ \\
                Liver & $0.05$ \\
                Skin & $0.01$ \\
                \end{tabular}
                \label{tab:eff}
            \end{ruledtabular}
        \end{table}

>>>>

%documentclass{article}
\documentclass[nofootinbib,preprint,aps]{revtex4-1}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{xspace}
\usepackage{csquotes}
\usepackage{gensymb}
\usepackage{graphicx}
\graphicspath{ {images/} }
\usepackage{physics}
\usepackage{color}
\newcommand{\red}[1]{\textcolor{red}{\bf #1}}
\begin{document}
\title{Nuclear Waste Management and Disposal}

@@

\date{\today}%
\begin{abstract}
%    As nuclear reactors around the world continuously generate highly radioactive waste,
%    how we manage nuclear waste is becoming more pertinent. In this work,
%    we examine a few scientific and technical issues related to nuclear waste management and disposal.
%    The principles of radioactivity and nuclear fission help us to understand the sources and properties
%    of different types of nuclear waste. Focusing on the highly radioactive spent fuel, the rest
%    of this work outlines the advantages and drawbacks of different spent fuel management and disposal
%    methods. We study the methods used for performance assessments for Yucca Mountain Nuclear Waste Repository,
%    especially in the context of volcanic disruption. Finally, we discuss some possible options for 
%    nuclear waste management and disposal after
%    the termination of the Yucca Mountain project.\\
%    Word count: 7449\\
\end{abstract}
\maketitle
\tableofcontents
\newpage
\section{Introduction}
Nuclear power continues to play an important role in global energy supply.
The International Atomic Energy Agency (IAEA) estimates that nuclear power
consumption will increase by $40\%$ by 2030.\cite{iaea12}
Whereas the ``front end'' of the nuclear fuel cycle, such as reactor physics and power output, draws
the most attention from the public, the ``back end'' of the fuel cycle, i.e. waste management and
disposal, deserves equal, if not more attention.
Currently, the IAEA estimates that there are over 10,000 metric tons of highly radioactive nuclear waste 
generated worldwide annually,
most of which is spent nuclear fuel from nuclear reactors.\cite{iaea08, r12}
The world will have over $457,000$ metric tons of highly radioactive waste stock by the end of 2020.\cite{r12}
Highly radioactive waste remains hazardous for up to $10^6$ years after it is produced and thus
requires extremely careful disposal methods.
Moreover, there is currently no permanent repository in the world to dispose of these highly radioactive
waste.\footnote{The Waste Isolation Pilot Plant (WIPP) in New Mexico is designed to be a permanent radioactive
    waste repository, but only
for waste classified as transuranic, which, by the Nuclear Regulatory Commission standard, does not have particularly
high radioactivity but only long half-life (See Sec.~\ref{sec:waste:classification}). In particular, WIPP does not accept spent nuclear
fuel.}

It is alarming that more than $75\%$ of the spent nuclear fuel worldwide is stored on-site at the nuclear power plants
in water pools that have exceeded their designed capacity and are reaching their limiting capacity.\cite{aa12}
Two weeks after the devastating earthquake and tsunami hit the Fukushima Daiichi Nuclear Power Plant in Japan,
the water pool cooling system was cut off from power. Officials feared that
the highly radioactive waste in the pools would catch fire and send radioactive smoke
across a much larger part of Japan, including Tokyo.\cite{s16} For some ``fortuitous'' reasons, the fear did not come true.
However, the Nuclear Regulatory Commission (NRC), the independent government agency
of the U.S. government overseeing nuclear energy safety,
estimates that a major fire in a U.S. spent fuel pool would
displace 3.4 million people in an area larger than New Jersey.\cite{s16}

@@

In this work, we will attempt to examine some of the scientific issues underlying nuclear waste management
and disposal. We will start by introducing radiation and its biological effects (Section~\ref{sec:phys}); then
we examine the sources and the physical properties of nuclear waste (Section~\ref{sec:waste}),
its management and temporary storage (Section~\ref{sec:temp}), and the debates around its eventual disposal (Section~\ref{sec:disposal}).

\section{Radioactivity and Its Biological Effects}
\label{sec:phys}
The principles of radioactivity and its biological effects play a central role in our
discussions of nuclear waste. Hence, we should first review and extend our understanding of 
radioactivity and radiation.
    \subsection{Radioactivity\footnote{
    The derivations in this section follow chapter $6$ in
    Krane's text \cite{k88} and chapter $1$ in Lilley's text \cite{l01}.
    }}
    Radioactive decay is the process in which a nucleus decays from a higher energy state to a lower
    energy state and releases energy in the form of radiation. Radioactivity accounts for
    most of the processes that generate detrimental radiation in nuclear waste.
    Here we review the physical principles and mathematical formalism behind it.
        \subsubsection{Quantum theory of radioactive decays}
        The law of radioactive decay is a direct consequence of quantum mechanics.
        Consider a nucleus going through radioactive decay.
        We can approximate the nucleus as a particle situated in a potential well with potential given by
        \begin{equation}
            \label{eq:vp}
            \tilde{V}(t) = V + V_p,
        \end{equation}
        where $V$ is the time-independent nucleus potential that gives rise to eigenstates $\ket{\psi_1}$
        and $\ket{\psi_2}$ with $\ket{\psi_2}$ having a lower (expected) energy; $V_p$ is a small
        perturbation that causes the decay.
        Suppose the particle is in a state given by
        \begin{equation}
            \label{eq:state}
            \ket{\psi(t)} = c_1(t)\ket{\psi_1} + c_2(t)\ket{\psi_2},
        \end{equation}
        where $c_1,c_2$ are the coefficients of superposition.
        A decay from $\ket{\psi_1}$ to $\ket{\psi_2}$ is equivalent to
        having $|c_2(t)|^2\gg |c_1(t)|^2$ as time increases. Since we know that time-independent potentials
        only give rise to time-independent states, the fact that Eq.~\ref{eq:state} has time dependence requires
        our perturbation $V_p$ in Eq.~\ref{eq:vp} to be time dependent.

@@

        Specific activity, on the other hand, gives the activity per mass of a sample
        \begin{equation}
        \mathcal{A}_s = \frac{\mathcal{A}(0)}{{m_{sample}}}.
        \end{equation}
        Specific activity can characterize the activity level of a certain isotope or compound.
        The difference between
        activity and specific activity resembles the difference between heat capacity and specific heat
        capacity.
        The radioactive isotope used for Computer Tomography (CT) scan has a specific activity
        of $\sim 0.01$ Ci/g. See Table~\ref{tab:hl} for the half-lives and specific activities
        of some isotopes present in highly radioactive nuclear waste.
        \begin{table}
            \centering
            \caption{Specific activity and half-life of isotopes commonly present in spent nuclear fuel \cite{toi}}
            \begin{ruledtabular}
                \begin{tabular}{l l l}
                Isotope & Half-life (Years) & Specific Activity (Ci/g)\\
                \hline
                Cesium-137 & $37$ & $88$\\
                Thorium-232 & $1.4 \times 10^{10}$ & $1.1\times 10^{-7}$\\
                Uranium-233 & $1.6\times 10^6$ & $9.3\times 10^{-3}$\\
                Plutonium-239 & $2.4\times 10^4$ & $0.063$\\
                Neptunium-237 & $2.1\times 10^6$ & $6.9\times 10^{-4}$\\
            \end{tabular}
            \label{tab:hl}
            \end{ruledtabular}
        \end{table}
        \subsubsection{Types of radiation}
        There are three main types of radioactive decay processes that generate radiation in nuclear waste: 
        Alpha decay, which generates alpha particles (helium-4 nuclei), beta minus decay which generates beta
        particles (electrons) and gamma decay which generates gamma rays.
        The types of radiation (particle beam) generated
        have different properties and deserve separate treatments when we discuss health effects and shielding.
        \paragraph{Alpha decay}
        An alpha particle is a positively-charged 
        helium-4 nucleus ${}^4_2$He$^{2+}$. In an alpha decay, a nucleus emits an alpha
        particle and decays into a different element. The emission of an alpha particle is
        particularly favorable because it has a particularly
        high binding energy. The process can be represented by
        \begin{equation}
            {}^A_Z X_N \rightarrow {}^{A-4}_{Z-2}X'_{N-2} + \alpha,
        \end{equation}
        where $Z$ denotes the atomic number (number of protons) of a nucleus $A$ the atomic mass, and $N$ the
        neutron count.

@@

        \paragraph{Gamma decay}
        Most nuclear reactions, including the alpha and beta decay, leave the resulting nucleus in
        an excited state.
        We know that excited atoms radiate photons to fall back into ground state. Similarly, excited nuclei
        decay to ground state via emitting photons, with energies in the range
        of $0.1$ to $10$ MeV, which are characteristic of the energy differences between nuclear energy states.
        The energy range correponds to that of a gamma ray photon.

        \subsubsection{Daughter nuclei and subsequent decay\footnote{The derivation in this section follows that of
        \cite[chapt. 1.5]{l01}}}
        The resultant nuclei (called ``daughter products'') from a radioactive decay are often radioactive
        themselves and their decay rate can be larger or smaller than their parents. Consider a decay chain
        $A\rightarrow B \rightarrow C \rightarrow ...$ with decay constants $\lambda_A, \lambda_B,...$. Suppose
        we start with a sample of $N_0$ $A$ nuclei. The change of the number of nuclei $B$ increases as $A$
        decays and decreases s $B$ decay. Hence we have
        \begin{equation}
            \frac{dN_B}{dt} = -\lambda_B N_B + \lambda_A N_A.
        \end{equation}
        Multiply both sides by an integrating factor $e^{\lambda_B t}$ and substitute in the number of particle
        $A$ as a function of time $N_A=N_0 e^{-\lambda_A t}$ to get
        \begin{align}
            \frac{dN_B}{dt}e^{\lambda_B t} + \lambda_B N_B e^{\lambda_B t} &= \lambda_A N_A e^{\lambda_B},\nonumber\\
            \frac{d}{dt} (N_B e^{\lambda_B t}) &= \lambda_A N_0 e^{(\lambda_B-\lambda_A)t}, \nonumber\\
            N_B e^{\lambda_B t} &= \lambda_A N_0 \int\limits_0^t e^{(\lambda_B - \lambda_A)u} du \nonumber\\
                                &= \frac{\lambda_A}{\lambda_B-\lambda_A}N_0 e^{(\lambda_B - \lambda_A)t}
        \end{align}
        Hence we end up with the number of $B$ nuclei as a function of time
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B-\lambda_A}N_0(e^{-\lambda_A t} - e^{-\lambda_B t}).
            \label{eq:daughter}
        \end{equation}
        Similar derivations can be done for the rest of the isotopes in the decay chain by considering
        one decay in a chain at a time. If the daughter isotope has a longer half-life than the parent
        ($\lambda_A \gg \lambda_B$), for small $t$, Eq.~\ref{eq:daughter} becomes
        $N_B(t)\sim N_0 e^{-\lambda_A t}$. After a long time,
        Eq.~\ref{eq:daughter} becomes $N_B(t) \sim N_0 e^{-\lambda_B t}$; the decay
        chain can be treated as the fast decay of $A$ into $B$, and then the slow decay of $B$
        into $C$.

        On the other hand, if the parent is longer lived
        ($\lambda_B \gg \lambda_A)$, after a long time, Eq.~\ref{eq:daughter} can be written as
        \begin{equation}
            N_B(t)=\frac{\lambda_A}{\lambda_B}N_0 e^{-\lambda_A t}=\frac{\lambda_A}{\lambda_B}N_A(t).
        \end{equation}
        Hence the number of nuclei $A$ and nuclei $B$ are in equilibrium determined by their decay rates, with
        $A$ dominating the nuclei count. Taking the limit as $\lambda_A \rightarrow 0$, $N_B(t)\rightarrow 0$.
        The decay is essentially the decay of $A$ directly into $C$. Hence, if we have a chain of reaction
        where each daughter nucleus has a half-life much shorter than that of its parent, then 
        isotope $A$ effectively decays into the end of the decay chain and releases a lot more radiation than
        in a single decay. Moreover, before the decay chain reaches its equilibrium, the activity of the sample
        may even increase.

        \subsection{Biological effects of radiation\footnote{Much of the discussion in this section follows that in
        Lilley's text \cite[chapt. 7]{l01}}}
        The biological effects of radiation have been an important topic for decades. A large
        number of studies took place after the two nuclear bomb explosions in Japan in 1945
        and after disastrous nuclear plant accidents like Chernobyl. Therefore, we have a relatively
        thorough understanding of the consequences of mass exposure to radiation. However, relatively
        little is known about the long-term effects of exposure to small doses, including subtle biological
        changes that lead to cancer and genetic defects. Here we briefly summarize the known effects.
        \subsubsection{Interactions of radiation with living tissues}
        \label{sec:interactions}
        Radiation causes damage to living tissues by depositing energy into biological molecules.
        Depending on the type of the radiation
        and the molecules present in the tissues, the damage mechanisms and effects are different.

        An alpha particle loses energy predominantly via the interactions between its electric
        field and the electrons in the matter it travels through. Interactions with nuclei and head-on collisions
        are quite rare. Since the speed of an alpha particle is low, it deposits more energy as it
        travels. In living tissues, an $1$-MeV alpha particle travels through only a few tens of micrometers and thus
        are easily stopped by skin. However, if alpha particles are ingested and more sensitive organs are irradiated,
        serious damages can be done.

        Beta radiation consistens of electrons. Electrons are much lighter and hence are easily scattered.
        Moreover, they are much more penetrating since their energy lost per distance traveled is lower.
        Finally, electrons also radiate photons
        via bremsstrahlung radiation as they decelerate. Therefore, the electrons' energy end up distributing
        across a much larger volume.

        Gamma rays interact with matter via photoelectric effects, Compton scattering and pair productions
        depending on the energy of the photon and the atoms it interact with.
        Since biological tissues mainly consist of elements of low atomic number, Compton
        scattering is the most important interaction. The scattered photons may interact further with the tissues
        via photoelectric effects and secondary Compton scattering. As a result, the intensity of the energy deposition falls
        off much more slowly and is hence more penetrating. Both electrons and photons can penetrate human body easily.
        Although they do not deposit as much energy as alpha particles, they are generally more dangerous due to their
        penetration abilities.

        The energy deposition described above is an important cause of irreversible damage to sensitive molecules
        like the Deoxyribonucleic (DNA) molecules if the energy transferred is large enough.
        In the cases where the energy transferred is not as large, the radiation ionizes simpler molecules like water
        to form chemically active free radicals. The free radicals have unpaired electrons. Hence they go on to react
        and disrupt
        other molecules that may be crucial to the cells.

        \subsubsection{Quantifying radiation on living tissues}
        The biological impact of radiation depends on the type, power and energy of the radiation, the
        exposure time, and the part of the body exposed. Various quantities was developed to
        describe the biological impacts of radiation. Here we review three
        of these quantities: the absorbed dose, the equivalent dose and the effective dose.

        The effects of radiation is usually characterized by the amount of energy it carries since it constrains
        the amount of energy it deposits in living tissues.
        Absorbed dose~($D$) quantifies the 
        radiation energy absorbed per unit mass. The SI unit for absorbed dose is gray (Gy), which is
        equivalent to $J/kg$.

        For an external $\gamma$-ray source emitting $\mathcal{A}$ million photons per second 
        of energy $E_{\gamma}$
        (in MeV), one can approximate the dose rate (dose received per unit of time) by \cite[chapt. 7]{l01}
        \begin{equation}
            \frac{dD}{dt}(\mu\text{Gy h}^{-1}) \approx \frac{\mathcal{A}\text{(MBq)}\times E_{\gamma}\text{(MeV)}}
            {6\times [r\text{(m)}]^2}.
        \end{equation}
        The approximate linear dependence on $E_{\gamma}$ is valid from less than $0.1$ MeV to several
        MeV and therefore should be good for all sources we are considering.\cite{my68}
        The inverse-squared dependence comes from the isotropic radiation assumption.

        \begin{table}
            \centering
            \caption{Weighting factors for different types of radiation \cite{icrp74}}
            \begin{ruledtabular}
                \begin{tabular}{l c c}
                Type of radiation & Energy range & Weighting factor, $w_R$\\
                \hline
                Photons, electrons & All energies & 1\\
                Neutrons & $<10$ keV & 5 \\
                         & $10-100$ keV & 10 \\
                         & $100$ keV-$2$ MeV & 20 \\
                         & $2-20$ MeV & 10 \\
                Protons & $<20$ MeV & 5 \\
                Alpha particles, fission fragments, heavy nuclei & & $20$\\
            \end{tabular}
            \label{tab:eq}
            \end{ruledtabular}
        \end{table}
        However, the effects of the radiation on living tissue depend on the type of radiation.
        For example, the same dose of alpha particle deposit more energy in living tissue than gamma rays.
        To quantify the {\it effects} of radiation, we introduce equivalent
        dose ($H$) as defined by 
        \begin{equation}
        H=w_R \times D_R,
        \end{equation}
        where $D_R$ is the absorbed dose and $w_R$ the weighting factor of a given kind of radiation $R$
        (see Table~\ref{tab:eq} for a list of weighing factors). The unit for equivalent dose is sievert (Sv).
        $1$ Sv is defined as the effects equivalent to $1$ Gy of Gamma radiation on living tissues. It describes
        the effects rather than the physical dose received by living tissues.

        Finally, certain organs and parts of the body are more vulnerable to radiation than the rest.
        Therefore, effective dose~($E$) is defined as the equivalent dose weighted by the body part
        radiated
        \begin{equation}
            E = w_T \times H_T,
        \end{equation}
        where $w_T$ is the body tissue weighting factor, and $H_T$ the equivalent dose on the tissue
        at question. Since effective dose quantifies the effects of radiation, the unit for effective dose
        is also sievert.
        Some of the recommended values of tissue
        weighting factors are given in Table~\ref{tab:eff}. A weighing factor of $1$ is equivalent to the
        whole body absorbing the radiation uniformly.
        \begin{table}
            \centering
            \caption{Weighting factors for individual organs \cite{icrp74}}
            \begin{ruledtabular}
                \begin{tabular}{l c}
                Tissue & Weighting factor, $w_T$\\
                \hline
                Gonads & $0.20$\\
                Red bone marrow & $0.12$ \\
                Liver & $0.05$ \\
                Skin & $0.01$ \\
                \end{tabular}
                \label{tab:eff}
            \end{ruledtabular}
        \end{table}
